### 2022-03-20 iwsm/abstract.tex
Proofread
Integrate review comments

\abstract{
Labelled networks are an important class of data,
naturally appearing
in numerous applications in science and engineering.
A typical inference goal is to determine how the vertex labels
(or {\em features}) affect the network's structure.
In this work, we introduce a new generative model, the feature-first block model (FFBM),
that facilitates the use of rich queries on labelled networks.
We develop a Bayesian framework and devise a two-level Markov chain Monte 
Carlo approach to efficiently sample from the
relevant posterior distribution of the FFBM parameters. This allows us to infer if and how the observed vertex-features affect macro-structure.
We apply the proposed methods to several real-world networks
to extract the most important features along which the vertices
are partitioned. The approach stands out from its peers in
that the whole feature-space is used automatically and features
can be rank-ordered implicitly by importance.
}

>>>>

\abstract{
Labelled networks are an important class of data,
naturally appearing
in applications in science and engineering.
A typical inference goal is to determine how the vertex labels
(or {\em features}) affect the network's structure.
We introduce a new generative model, the feature-first block model (FFBM),
that facilitates the use of rich queries on labelled networks.
We develop a Bayesian framework and devise a two-level Markov chain Monte 
Carlo approach to efficiently sample from the
posterior distribution of the FFBM parameters. This allows us to infer if and how the observed vertex-features affect macro-structure.
We apply the proposed methods to several real-world networks
to extract the most important features along which the vertices
are partitioned. Importantly, the whole feature-space is used automatically and features
can be rank-ordered implicitly by importance.
[The full version of this paper is available on arXiv as [cs.LG] 2105.13762.]
}

### 2022-03-20 iwsm/experiments.tex
Refactor
Integrate review comments

We apply our proposed methods to a variety of real-world datasets.
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
%We employ the following metrics to assess model performance. 
%First, the average
%description length per entity (nodes and edges) 
%$\bar{S}_e$ 
%used to gauge the SBM fit is defined as:
%%
%\begin{equation}
%	\bar{S}_e \triangleq \frac{1}{(N+E) |\mathcal{T}_b|} \sum_{t\in \mathcal{T}_b} S \left( b^{(t)} \right).
%	\label{eqn:mean-dl}
%\end{equation}
%%
%Next, to assess the performance of the feature-to-block predictor, 
%the vertex set $[N]$ 
%is partitioned at random so that 
%a constant fraction $f$ of vertices form the training set $\mathcal{G}_0$ and 
%the remainder form the test set $\mathcal{G}_1$.
%The $b$-chain is run using the whole network but only vertices $v \in \mathcal{G}_0$ are 
%used for the $\theta$-chain. 
%Then the  the average cross-entropy loss 
%over each set is used to gauge the quality of the fit,
%%
%\begin{equation}
%	\bar{\mathcal{L}}_\star \triangleq \frac{1}{|\mathcal{T}_\theta|} \sum_{t \in \mathcal{T}_\theta} \mathcal{L}_\star^{(t)},
%	\quad \textrm{where} \quad
%	\mathcal{L}_\star^{(t)} \triangleq \frac{1}{|\mathcal{G}_\star|} \sum_{i \in \mathcal{G}_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
%	\label{eqn:cross-entropy-loss}
%\end{equation}
%%
%where $\star \in \{0, 1\}$ toggles between the training and test sets
%and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%%
%Nevertheless, the cross-entropy loss is a coarse measure of fit. 
%A new measure, specific to each detected block,
%can be defined as follows. Let
%$\mathcal{B}_\star(j)$ 
%be the set of vertices with maximum a posteriori probability of belonging 
%to block~$j$,
%$
%	\mathcal{B}_\star(j) \triangleq \{i \in \mathcal{G}_\star : \hat{b}_i = j\},
%$
%where
%$ 
%	\hat{b}_i \triangleq \underset{j}{\operatorname{argmax}}\hat{y}_{ij},
%$
%and
%define the {\em block-accuracy} for block $j$ as,
%%
%\begin{equation}
%	\eta_\star(j) \triangleq \frac{1}{|\mathcal{B}_\star (j)| \cdot 
%	|\mathcal{T}_\theta| } 
%	\sum_{i \in \mathcal{B}_\star (j)}  \sum_{t \in \mathcal{T}_\theta}
%	\boldsymbol{1} \left\{\hat{b}_i = \underset{j}{\operatorname{argmax}}\phi_j \left( x_i; \theta^{(t)} \right) \right\}.
%	\label{eqn:accuracy}
%\end{equation}
%%
%This effectively tests whether the feature-to-block and 
%graph-to-block predictions agree in their largest component.
%For the higher-dimensional datasets, we also apply the 
%dimensionality reduction method 
%of Section~\ref{sec:dim-reduction}.  
%We then retrain the feature-block predictor using only the retained 
%feature set $\mathcal{D}'$, and report the log-loss over the training and 
%test sets for the reduced classifier -- 
%denoted $\bar{\mathcal{L}}_0'$ and $\bar{\mathcal{L}}_1'$ respectively.

>>>>

We apply our proposed methods to a variety of real-world datasets.
The inferred partitions $b$ for all of these are given on Figure~\ref{fig:graphs-all}.
To assess model performance,
%
the average
description length per entity (nodes and edges) 
$\bar{S}_e$ 
is used to gauge the SBM fit,
and
the vertex set $[N]$ is partitioned at random into training and test sets, 
$\mathcal{G}_0$ and $\mathcal{G}_1$,
%
to assess the performance of the feature-to-block predictor.
The average cross-entropy loss 
over each set,
denoted $\bar{\mathcal{L}}_\star$,
is used to gauge the quality of the fit.

For higher-dimensional datasets, we develop a novel 
dimensionality reduction method to select only the top $D'$ features.  
We then retrain the feature-block predictor using only the retained 
feature set, and report the cross-entropy loss 
$\bar{\mathcal{L}}_\star'$
over the training and 
test sets for the reduced classifier.

Table~\ref{tab:results} summarises the results for each experiment. 
We see that the dimensionality reduction procedure 
brings the training and test losses closer, indicating that
the retained features
are indeed well correlated with the underlying graphical 
partition and that the approach generalises correctly.

### 2022-03-20 iwsm/ffbm.tex
Integrate review comments

We propose a novel generative model for labelled networks -- which we call the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}.
Let $N$ denote the number of vertices, $B$ the number of blocks
and $D$ the number of features associated with each vertex.
We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.
%
For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$ -- noting the notation $[K] = \{1, 2 \dots K\}$. For each vertex $i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector $x_i$, independently between vertices, 
$p(b| X, \theta) = \prod_{i \in [N]} \phi_{b_i} (x_i; \theta)$.
%
\begin{figure}[!ht]
	\centering
%   IWSM Template does not support TIKZ
%	\begin{tikzpicture}[
%		roundnode/.style={circle, draw=black, minimum size=7mm},
%		squarednode/.style={rectangle, draw=black, minimum size=7mm}
%		]
%		% nodes
%		\node[roundnode] (X) at (0, 0) {$X$};
%		\node[squarednode] (b) at (3, 0) {$b$};
%		\node[roundnode] (A) at (6, 0) {$A$};
%		
%		% arrows
%		\draw[->] (X.east) -- node[above] {$\theta$} (b.west);
%		\draw[->] (b.east) -- node[above] {$\psi$}(A.west);
%	\end{tikzpicture}
	\includegraphics[width=0.8\textwidth]{img/ffbm.png}
	\caption{The Feature-First Block Model (FFBM)}
	\label{fig:ffbm}
\end{figure}

Once the block memberships $b$ have been generated, we then draw the 
adjacency matrix of the graph, $A$, from the microcanonical DC-SBM, Peixoto (2017), with additional parameters 
$\psi$,
$
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi).
	\label{eqn:A-generation}
$
Priors on the parameters $\theta$ and $\psi$ are chosen to complete the Bayesian framework.

>>>>

We propose a novel generative model for labelled networks, which we call the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}.
Let $N$ denote the number of vertices, $B$ the number of blocks
and $D$ the number of features associated with each vertex.
We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.
%
For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$, where we write $[K] = \{1, 2 \dots K\}$. For each vertex~$i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector~$x_i$, independently between vertices, so
$p(b| X, \theta) = \prod_{i \in [N]} \phi_{b_i} (x_i; \theta)$.
%
\begin{figure}[!ht]
	\centering
%   IWSM Template does not support TIKZ
%	\begin{tikzpicture}[
%		roundnode/.style={circle, draw=black, minimum size=7mm},
%		squarednode/.style={rectangle, draw=black, minimum size=7mm}
%		]
%		% nodes
%		\node[roundnode] (X) at (0, 0) {$X$};
%		\node[squarednode] (b) at (3, 0) {$b$};
%		\node[roundnode] (A) at (6, 0) {$A$};
%		
%		% arrows
%		\draw[->] (X.east) -- node[above] {$\theta$} (b.west);
%		\draw[->] (b.east) -- node[above] {$\psi$}(A.west);
%	\end{tikzpicture}
	\includegraphics[width=0.6\textwidth]{img/ffbm.png}
	\caption{The Feature-First Block Model (FFBM)}
	\label{fig:ffbm}
\end{figure}

Once the block memberships $b$ have been generated, we then draw the 
adjacency matrix of the graph 
	$A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi)$
from the microcanonical DC-SBM, Peixoto (2017), with additional parameters 
$\psi$.
Appropriate priors are placed on the parameters $\theta$ and $\psi$ to 
complete the Bayesian framework.

### 2022-03-20 iwsm/inference.tex
Integrate review comments

Given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution for $\theta$,
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
Therefore, following standard Bayesian practice,
we aim to draw samples from this posterior,
$
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X).
$
We propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain these samples
$\{\theta^{(t)}\}$. We first draw a sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$:
%
\begin{equation}
	b^{(t)} \stackrel{\rm distr}{\approx} p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)} \stackrel{\rm distr}{\approx} 
	p\big(\theta | X, b^{(t)} \big),
\end{equation}
%
where these approximations become exact as
the number of MCMC iterations $t\to\infty$.
Splitting the Markov chain into two levels side-steps the intractable summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior.
%
\begin{figure}[!ht]
	\centering
%   IWSM Template does not support tikz
%	\begin{tikzpicture}[
%		scale=0.8, every node/.style={transform shape},
%		roundnode/.style={circle, draw=black, minimum size=12mm},
%		squarednode/.style={rectangle, draw=black, minimum size=12mm}
%		]
%		% nodes
%		\node[roundnode] (b0) at (0, 2) {$b^{(t)}$};
%		\node[roundnode] (b1) at (5, 2) {$b'$};
%		\node[roundnode] (b2) at (10, 2) {$b^{(t+1)}$};
%		\node[roundnode] (t0) at (0, 0) {$\theta^{(t)}$};
%		\node[roundnode] (t1) at (5, 0) {$\theta'$};
%		\node[roundnode] (t2) at (10, 0) {$\theta^{(t+1)}$};
%		
%		% arrows
%		\draw[->] (b0) to node[above] {$\sim q_b ( b^{(t)}, b' | A )$} (b1);
%		\draw[->] (b1) to node[above] {$\alpha_b (b, b' | A )$} (b2);
%		\draw[->] (b0) [out=25, in=155] to node[above] {$1-\alpha_b$} (b2);
%		
%		\draw[->] (t0) to node[above] {$\sim q_\theta(\theta^{(t)}, \theta' | X, b^{(t)})$} (t1);
%		\draw[->] (t1) to node[above] {$\alpha_\theta (\theta, \theta' | X, b^{(t)})$} (t2);
%		\draw[->] (t0) [out=-25, in=-155] to node[below] {$1-\alpha_\theta$} (t2);
%		
%		\draw[dashed, ->] (b0) to (2, 0.8);
%		\draw[dashed, ->] (b0) to (6.5, 0.8);
%		
%	\end{tikzpicture}
	\includegraphics[width=0.6\textwidth]{img/sampling-sequence}
	\caption{$\theta$-sample generation.}
	\label{fig:samp-sequence}
\end{figure}

>>>>

Given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution for $\theta$,
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
Following standard Bayesian practice,
we propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain 
samples $\theta^{(t)}$ from this posterior.
First we sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$,
%
\begin{equation}
	b^{(t)} \sim p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)}\sim
	p\big(\theta | X, b^{(t)} \big).
\end{equation}
%
% where these approximations become exact as
% the number of MCMC iterations $t\to\infty$.
Splitting the Markov chain into two levels side-steps the intractable summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior.
%
\begin{figure}[!ht]
	\centering
%   IWSM Template does not support tikz
%	\begin{tikzpicture}[
%		scale=0.8, every node/.style={transform shape},
%		roundnode/.style={circle, draw=black, minimum size=12mm},
%		squarednode/.style={rectangle, draw=black, minimum size=12mm}
%		]
%		% nodes
%		\node[roundnode] (b0) at (0, 2) {$b^{(t)}$};
%		\node[roundnode] (b1) at (5, 2) {$b'$};
%		\node[roundnode] (b2) at (10, 2) {$b^{(t+1)}$};
%		\node[roundnode] (t0) at (0, 0) {$\theta^{(t)}$};
%		\node[roundnode] (t1) at (5, 0) {$\theta'$};
%		\node[roundnode] (t2) at (10, 0) {$\theta^{(t+1)}$};
%		
%		% arrows
%		\draw[->] (b0) to node[above] {$\sim q_b ( b^{(t)}, b' | A )$} (b1);
%		\draw[->] (b1) to node[above] {$\alpha_b (b, b' | A )$} (b2);
%		\draw[->] (b0) [out=25, in=155] to node[above] {$1-\alpha_b$} (b2);
%		
%		\draw[->] (t0) to node[above] {$\sim q_\theta(\theta^{(t)}, \theta' | X, b^{(t)})$} (t1);
%		\draw[->] (t1) to node[above] {$\alpha_\theta (\theta, \theta' | X, b^{(t)})$} (t2);
%		\draw[->] (t0) [out=-25, in=-155] to node[below] {$1-\alpha_\theta$} (t2);
%		
%		\draw[dashed, ->] (b0) to (2, 0.8);
%		\draw[dashed, ->] (b0) to (6.5, 0.8);
%		
%	\end{tikzpicture}
	\includegraphics[width=0.6\textwidth]{img/sampling-sequence}
	\vspace*{-0.1in}
	\caption{$\theta$-sample generation.}
	\label{fig:samp-sequence}
\end{figure}

### 2022-03-20 iwsm/introduction.tex
Refactor
Proofread
Integrate review comments

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
%Finding ways to recover the latent communities from the observed graph is an important
%task in many applications,
%including
%compression [Abbe (2016)] and link prediction 
%[Zhu et al. (2016)].
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model;
the standard is the stochastic block model (SBM) -- see Peixoto (2017).
%Numerous variants of the SBM  have been proposed, e.g., the MMSBM -- Airoldi et al. (2009) --
%and OSBM -- Zhu et al. (2016) -- but these do not include features in the graph generation
%process.

Analysing a labelled network with a simple SBM variant, requires partitioning the graph into blocks grouped by distinct values of the feature of interest. The associated model can then be used to test for evidence of heterogeneous connectivity between the feature-grouped blocks. Nevertheless, this approach can only consider disjoint feature sets and the feature-grouped blocks are often an unnatural partition of the graph.

We would instead prefer to partition the graph into its most natural blocks and then find which of the available features -- if any -- best predict the resulting partition. Thus motivated, we present a novel framework for modelling labelled networks.
This would not the first extension of the SBM to labelled networks, e.g. Stanley et al (2019). However, most of the current approaches are focused on leveraging feature information to partition the graph more reliably in the presence of noise.
We seek instead to develop a model well suited for inferring how vertex features impact graphical structure and to report our confidence in those conclusions.

>>>>

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model;
the standard is the stochastic block model (SBM), see Peixoto (2017).

Analysing a labelled network with one of the standard SBM variants requires partitioning the graph into blocks grouped by distinct values of the feature of interest. The associated model can then be used to test for evidence of heterogeneous connectivity between the feature-grouped blocks. But this approach can only consider disjoint feature sets and the feature-grouped blocks often provide an unnatural partition.

We would instead prefer to partition the graph into its most natural blocks and then find which of the available features -- if any -- best predict the resulting partition. Thus motivated, we present a novel framework for modelling labelled networks.
This is not the first extension of the SBM to labelled networks, e.g. Stanley et al (2019). However, most of the current approaches are focused on leveraging feature information to partition the graph more reliably in the presence of noise.
We seek instead to develop a model well suited for inferring how vertex features impact graphical structure and to report our confidence in those conclusions.
